---
title: 'Kelleher 2015: Chapter 4, Exercise 9'
author: "Hamilton Hoxie Ackerman"
date: "3/6/2017"
output: html_document
---

In this problem, we're calculating the probability that a model ensemble that uses simple majority voting will make an incorrect prediction if...

#### 9a)

11 independent models, all of which have an error rate of 0.2

Let $X = $ the number of models that make an incorrect decision. Then $X \sim Bin(11, 0.2)$ and we make an incorrect decision iff $X \gt 5$.

```{r}
pbinom(5, size=11, prob=0.2, lower.tail=FALSE)
```

Pretty unlikely, as you would expect with such a small error rate. (These are much stronger than weak learners, for example.)

#### 9b)

11 independent models, all of which have an error rate of 0.49.

Let $X = $ the number of models that make an incorrect decision. Then $X \sim Bin(11, 0.49)$ and we make an incorrect decision iff $X \gt 5$.

```{r}
pbinom(5, size=11, prob=0.49, lower.tail=FALSE)
```

Just less than half, which seems reasonable. As we increase our sample size, this probability should decrease, as the standard error decreases like $\sqrt(n)$...

#### 9c)

21 independent models, all of which have an error rate of 0.49.

Let $X = $ the number of models that make an incorrect decision. Then $X \sim Bin(21, 0.49)$ and we make an incorrect decision iff $X \gt 10$.

```{r}
pbinom(10, size=21, prob=0.49, lower.tail=FALSE)
```

The probability didn't decrease much, but $n=21$ isn't much bigger than $n=11$; certainly not large enough for the CLT to kick in, for example.