---
title: 'Kelleher 2015: Chapter 4, Exercise 2'
author: "Hamilton Hoxie Ackerman"
date: "3/4/2017"
output: html_document
---

### 2a)

I could do this by hand, but letting the computer do the work seemed like a better idea. Finding a plain ID3 implementation was harder than it should have been, but I guess that fancier things are used in practice. Here's one I found (via http://r.789695.n4.nabble.com/Is-there-an-ID3-implementation-in-R-td4696375.html):

```{r, message=FALSE}
# Enter the data manually
ex2data <- data.frame(GoodBehavior = c(rep(FALSE, 3), rep(TRUE, 3)))
ex2data$AgeLT30 <- c(TRUE, FALSE, TRUE, rep(FALSE, 3))
ex2data$Drug <- c(rep(FALSE, 4), TRUE, FALSE)
ex2data$Recidivist <- rep(c(TRUE, FALSE), 3)

# Get ID3 classifier
#install.packages("RWeka")
library(RWeka)
#WPM("refresh-cache") 
#WPM("list-packages", "available")
#WPM("install-package", "simpleEducationalLearningSchemes") 
WPM("load-package", "simpleEducationalLearningSchemes") 
ID3 <- make_Weka_classifier("weka/classifiers/trees/Id3") 
```

```{r}
ID3(Recidivist ~ ., data = ex2data)
```

So...

* If Age <30, then Recidivist is predicted to be true
* If Age >= 30, then the Drug value provides the final prediction
* GoodBehavior isn't needed at all

### 2b)

Since the age isn't less than 30, DrugDependent gives us our predicted value of TRUE.

### 2c)

Since the age is less than 30, our predicted value is TRUE.
